The Effects of the NBA COVID Bubble on the NBA Playoffs

### 2022-02-13 Manuscript/NBABubbleAnalysis.Rmd
Update NBABubbleAnalysis.Rmd

# Introduction {#sec:intro}
Home-court advantage is often discussed in sports circles as a contributing
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge while playing at their homecourt, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a
bubble due to the COVID-19 pandemic
brought a great deal of concern for fans, teams, journalists, and others. The bubble environment
would not be able to replicate many of the factors that normal home-court advantage relies on.
Among the biggest is the effect from the crowd. The home crowd has the ability
to energize home teams, help control momentum in favor of the home team and 
create a chaotic and stressful environment for away teams to play in. There 
is no way to replicate these effects in the bubble conditions. Also, there are 
geographic factors that are lost, like altitude and time-zone effects. There's 
evidence across all sports that teams in high altitude regions can rely on away
teams coming in and struggling with lower oxygen levels at a higher altitude, 
which the home teams are accustomed to playing with \citet{Lopez}. Examples
of high altitude teams in the NBA would be the Utah Jazz and Denver Nuggets, 
both of whom participated in the 2020 NBA bubble. Also, referees are known to
be subject to favoring home teams. This is due in large part to pressure from fans, which
is no longer in play in the bubble. Refs may still have biases, but there's
nothing to sway that bias towards the home team like normal. Any positive 
effects of the home team playing in the arena they're most comfortable in 
and living in the comfort of their own home are also lost. Travel is also 
normally discussed as a factor hurting away teams, but in the context of
the playoffs in the bubble, losing this may not be significant. In the
playoffs, many times (aside from the first game of the series when the 
away team travels), both teams often travel at the same time, since they're 
both going back and forth between the two cities to complete the series.

@@

# Literature Review {#sec:litrev}
There is voluminous literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. @Kotecki reported significant evidence of home-court advantage by 
comparing field goal percentage, free throw percentage
and points scored in home vs. away teams. He found all of these measures showed evidence that
home-court advantage helps home teams play better. @Cao studied the effects
of pressure on performance in the NBA. Using free throws as their measure of
interest, they tested whether home
fans could distract and put pressure on opposing players to make free
throws. However, they did not find significant evidence
that home status has a substantial impact on missing from the free throw
line. @Harris used two-point shots, three-point shots and free throws as measures
of interest to study home-court advantage. Two-point shots were found to be
the strongest predictor of home-court advantage. They suggested
that home teams should try to shoot more two-point shots and force
their opponent to take more two-point shot attempts. This will 
help home teams have a greater control of the game play and help maximize the benefits 
home-court advantage.

Some studies focus less on shooting and more on other metrics.
For example, @Greer focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men’s basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance.  @Harville studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulated their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs. when playing at neutral court. This study
concluded with evidence supporting home-court advantage.

There are also surveys on the factors contributing to home-court advantage.
@Carron1992 gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site, even
if one team was designated as “home team”. This study was reviewed a decade later
by @Carron2005. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992, they have found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games across all major sports. There is less
evidence suggesting learning and rule factors impact across the various collegiate 
and professional sports. The NBA is not a league which has rules that
may favor the home team, like batting last in the MLB, but these rule factors 
also account for referee bias which may impact the NBA. One interesting finding cited
by @Carron2005 is that the absence of crowds results in overall performance increases.
Lastly, as discussed in the introduction, @Lopez find evidence that geographic
factors like altitude may influence and strengthen home-court advantage for teams in
high altitude regions.

>>>>

# Introduction {#sec:intro}
Home-court advantage is often discussed in sports circles as a contributing
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge while playing at their homecourt, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a
bubble due to the COVID-19 pandemic
brought a great deal of concern for fans, teams, journalists, and others. The bubble environment
would not be able to replicate many of the factors that normal home-court advantage relies on.
Among the biggest is the effect from the crowd. The home crowd has the ability
to energize home teams, help control momentum in favor of the home team and 
create a chaotic and stressful environment for away teams to play in. There 
is no way to replicate these effects in the bubble conditions. Also, there are 
geographic factors that are lost, like altitude and time-zone effects. There's 
evidence across all sports that teams in high altitude regions can rely on away
teams coming in and struggling with lower oxygen levels at a higher altitude, 
which the home teams are accustomed to playing with @Lopez. Examples
of high altitude teams in the NBA would be the Utah Jazz and Denver Nuggets, 
both of whom participated in the 2020 NBA bubble. Also, referees are known to
be subject to favoring home teams. This is due in large part to pressure from fans, which
is no longer in play in the bubble. Refs may still have biases, but there's
nothing to sway that bias towards the home team like normal. Any positive 
effects of the home team playing in the arena they're most comfortable in 
and living in the comfort of their own home are also lost. Travel is also 
normally discussed as a factor hurting away teams, but in the context of
the playoffs in the bubble, losing this may not be significant. In the
playoffs, many times (aside from the first game of the series when the 
away team travels), both teams often travel at the same time, since they're 
both going back and forth between the two cities to complete the series.

@@

# Literature Review {#sec:litrev}
There is voluminous literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. @Kotecki reported significant evidence of home-court
advantage by comparing field goal
percentage, free throw percentage and points scored in home vs.~away teams.
He found all of these measures showed evidence that
home-court advantage helps home teams play better. @Cao studied the effects
of pressure on performance in the NBA. Using free throws as their measure of interest, 
they tested whether home
fans could distract and put pressure on opposing players to make free
throws. However, they did not find significant evidence
that home status has a substantial impact on missing from the free throw line. 
@Harris used two-point shots, three-point shots and free
throws as measures of interest to study home-court advantage. Two-point
shots were found to be the strongest predictor of home-court advantage. They
suggested that home teams should try to shoot more two-point shots and force
their opponent to take more two-point shot attempts. This help home teams have a
greater control of the game play and help maximize the benefits 
home-court advantage.

Some studies focus less on shooting and more on other metrics.
For example, @Greer focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men's basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance. Another study focused
on scoring was conducted by @Jones, he analyzes scoring patterns across
each of the 4 quarters to analyze the effects of home-court advantage over
the course of a game. He found that typically 2/3 of the benefits of 
home-court advantage are received in the first quarter with the 
remaining 1/3 slowly accumulating over the final 3 quarters. This implies
home-court advantage is most effective when teams use it to build a 
lead early and maintain it for the remainder of the game. If they are 
losing after the first quarter, there is a sharp drop in win probability.

There are also surveys on the factors contributing to home-court advantage.
@Carron1992 gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site, even
if one team was designated as ``home team''. This study was reviewed a decade later
by @Carron2005. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992, they have found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games across all major sports. There is less
evidence suggesting learning and rule factors impact across the various collegiate 
and professional sports. The NBA is not a league which has rules that
may favor the home team, like batting last in the MLB, but these rule
factors also account for referee bias which may impact the NBA.  One
interesting finding cited by @Carron2005 is that the absence of crowds results
in overall performance increases. Another study by @Price is able 
to find some evidence of referee bias. They focus on two measures, DTOs and NTOs.
DTOs, discretionary turnovers, are defined as turnovers always caused by the ref
blowing the whistle while the ball is in play. NTOs, non-discretionary turnovers, 
are determined directly by players with no ref whistle, or the ball going out of
bounds. The use these to test ref bias by checking how variables, like home vs. away
team, affect DTOs relative to NTOs. They found evidence that a home bias does exist. 
In fact, home bias increases both during the playoffs and in games with higher attendance.
This is crucial to the NBA bubble, which consists of playoff games with no attendance. 
Lastly, as discussed in the introduction, @Lopez
find evidence that geographic factors like altitude may influence and strengthen home-court
advantage for teams in high altitude regions.

There are a few examples of natural experiments in basketball. @Harville studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulated their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs.~when playing on a neutral court. This study
concluded with evidence supporting home-court advantage. Also, @Boudreaux 
is able to construct a natural experiment using the Los Angeles Lakers and 
Los Angeles Clippers. Since these two teams share a home stadium, many factors like travel 
and familiarity are nullified. However, the designated home team has larger crowd
support due to attendance from their season ticket holders. By comparing
Lakers--Clippers matchups, they single out the effect of having a sympathetic
crowd and @Boudreaux estimate crowd effects increase the chance
of winning 21--22.8 percentage points.
